---
title: "PaperAlcohol_CrossSectionalSample.R"
author: "L. Cecilia López Steinmetz"
date: "17/2/2023"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

# R Code for:


# López Steinmetz LC, Leyes CA, Fong SB & Godoy JC. "Alcohol consumption, alcohol expectancies, and drinking contexts in young Argentinean college students before and during the COVID-19 pandemic: A one-year follow-up study"


### ABBREVIATIONS 


**Abbreviations:**
EA = Alcohol Expectancy (Expectativas hacia el alcohol)
CC = Consumption Contexts (Contextos de consumo de alcohol)


### LOAD THE DATASETS: b) Cross-sectional


#### b) Load the dataset: *CROSS-SECTIONAL sample*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
transv<-read.table("clipboard",header=TRUE, encoding = "Latin 1", dec=",", sep="\t") # CROSS-SECTIONAL
```


#### To explore the data


#### b) *CROSS-SECTIONAL sample*


*To order categories*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
transv$tsexo <- factor(transv$tsexo, levels=c("Wom", "Man")) 
```


*Check the type of the variables (CROSS-SECTIONAL) and transform them (when necessary)*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
str(transv)

transv$tunidadesalc<-as.numeric(transv$tunidadesalc)
transv$tALCSEM<-as.numeric(transv$tALCSEM)

str(transv)
```


### METHODS SECTION 


#### SKEWNESS, KURTOSIS, & MULTICOLLINEARITY


*To test Skewness and Kurtosis: See codes for supplementary files (at the end of these scripts)*


*To test for multicollinearity: See codes supplementary files (at the end of these scripts)*


#### DESCRIPTION OF THE SAMPLE


##### CROSS-SECTIONAL

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# Sex
table(transv$tsexo)

prop.table(table(transv$tsexo))*100


# Age
mean(transv$tedad) # at the first measurement

sd(transv$tedad) # at the first measurement


# Age of onset of alcohol consumption
mean(transv$tedadinicioalc)

sd(transv$tedadinicioalc)


# Weekly alcohol consumption (Units of alcohol)
mean(transv$tALCSEM)

sd(transv$tALCSEM)
```


##### CROSS-SECTIONAL 


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
#Load the dataset (CROSS-SECTIONAL)

# transv<-read.table("clipboard",header=TRUE, encoding = "Latin 1", dec=",", sep="\t") # CROSS-SECTIONAL

# transv$tsexo <- factor(transv$tsexo, levels=c("Wom", "Man")) # to order categories
```


### METHODS SECTION 


##### CROSS-SECTIONAL 

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
transvlog <- transv
```


##### LOG TRANSFORMATIONS

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES
transvlog$logtALCSEM <- log(transvlog$tALCSEM + 1)

# ALCOHOL EXPECTANCIES

# POSITIVE EA:
transvlog$logtEAPOS <- log(transvlog$tEAPOS + 1)

# NEGATIVE EA:
transvlog$logtEANEG <- log(transvlog$tEANEG + 1)

# CONSUMPTION CONTEXTS

# SOCIAL FACILITATION:
transvlog$logtfacilitsoc <- log(transvlog$tfacilitsoc + 1)

# PEER GROUP ACCEPTANCE:
transvlog$logtaceptpares <- log(transvlog$taceptpares + 1)

# PARENTAL CONTROL:
transvlog$logtctrolparent <- log(transvlog$tctrolparent + 1)

# STRESS CONTROL:
transvlog$logtctrolestres <- log(transvlog$tctrolestres + 1) 
```


*Preparing the data: based on LOG TRANSFORMED VALUES*


*To see the NAs*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(mice)
datatransv<-subset(transvlog,select=c(tsexo, tedad, tedadinicioalc, logtALCSEM, logtEAPOS, logtEANEG, logtfacilitsoc, logtaceptpares, logtctrolparent, logtctrolestres))
md.pattern(datatransv, rotate.names = TRUE)
md.pairs(datatransv) 
```


*We used the pmm (predictive mean matching) method for imputation*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# Libraries
# library(mice)
library(VIM)

# Missing data
p <- function(x) {sum(is.na(x))/length(x)*100}
apply(datatransv, 2, p) 
```


*Impute*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
set.seed(123)
```


*We run the mice code with 0 iterations*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
imptr <- mice(datatransv, maxit=0, seed = 123)
```


*Extract predictorMatrix and methods of imputation*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
predM <- imptr$predictorMatrix
meth <- imptr$method
```


*To view the first few rows of the predictor matrix*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
head(predM)
```


*See the methods of imputations*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
meth

imputeT <- mice(datatransv, m = 10, maxit = 5, predictorMatrix = predM, method = meth, seed = 123)

print(imputeT) 
imputeT$imp$logtEAPOS
imputeT$imp$logtEANEG 
imputeT$imp$logtfacilitsoc 
imputeT$imp$logtaceptpares  
imputeT$imp$logtctrolparent  
imputeT$imp$logtctrolestres 
```


*Distribution of oberserved/imputed values*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
xyplot(imputeT, logtEAPOS ~ logtALCSEM | .imp, pch = 20, cex=1.4)
xyplot(imputeT, logtEANEG ~ logtALCSEM | .imp, pch = 20, cex=1.4)
xyplot(imputeT, logtfacilitsoc ~ logtALCSEM | .imp, pch = 20, cex=1.4)
xyplot(imputeT, logtaceptpares ~ logtALCSEM | .imp, pch = 20, cex=1.4)
xyplot(imputeT, logtctrolparent ~ logtALCSEM | .imp, pch = 20, cex=1.4)
xyplot(imputeT, logtctrolestres ~ logtALCSEM | .imp, pch = 20, cex=1.4)
```


*Complete the dataset*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
newtransv <- complete(imputeT, 10) # We choosed the first imputation to complete the dataset

detach("package:mice", unload = TRUE)
detach("package:VIM", unload = TRUE)
```


*OUTCOME, EA AND CC MEASURED DURING QUARANTINE DUE TO THE COVID-19 PANDEMIC)*
*(CROSS-SECTIONAL SAMPLE)*


*To select the order in which the predictors will be included into the model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(leaps)

bestcrosssect<-regsubsets(logtALCSEM ~ tsexo + tedad + tedadinicioalc + logtEAPOS + logtEANEG + logtfacilitsoc + logtaceptpares + logtctrolparent + logtctrolestres, data = newtransv, nvmax = 11, method = "exhaustive")

summary(bestcrosssect, all.best = TRUE)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
plot(bestcrosssect, scale = "r2", main = "Best R^2") # R2 = 0.46
plot(bestcrosssect, scale = "adjr2", main = "Best Adjusted R^2") # adj R2 = 0.44
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# Model:
fullT<-lm(logtALCSEM ~ logtctrolestres + logtctrolparent + logtEAPOS + tedad + tsexo + logtfacilitsoc + logtEANEG + tedadinicioalc + logtaceptpares, data = newtransv)

summary(fullT)
```


*95% Confidence interval of best-fitted model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
confint(fullT, level = 0.95)
```


*ERROR RATE of best-fitted model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
sigma(fullT)/mean(newtransv$logtALCSEM)
# In our multiple regression, the Residual Standard Error (RSE) or sigma is 0.6705 corresponding to 35.57% error rate.
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
par(mfrow=c(2,2))
# Figure 
plot(fullT)
par(mfrow=c(1,1))
```


*Effect sizes*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
effectsize::cohens_f(fullT, ci = 0.95)
```




### SUPPLEMENTARY MATERIALS


#### To test Skewness and Kurtosis

#### & log transformation


*Testing Skewness and Kurtosis*


*Criteria:* range of acceptable values -1 to +1 for Skewness and -3 to +3 for Kurtosis (Brown, 2006) or near to.


*Reference:* Brown, T. A. (2006). Confirmatory factor analysis for applied research. New York: Guilford Press.


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(moments)
```


We have tested skewness and kurtosis in the original data (from which included NAs).
Then, we have tested skewness and kurtosis transforming the data by using the log transformation (natural logarithm).


#### CROSS-SECTIONAL SAMPLE #### 

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES
skewness(transv$tALCSEM)

kurtosis(transv$tALCSEM)

hist(transv$tALCSEM)

skewness(transvlog$logtALCSEM)

kurtosis(transvlog$logtALCSEM)

hist(transvlog$logtALCSEM)

# ALCOHOL EXPECTANCIES

# POSITIVE EA:
skewness(transv$tEAPOS, na.rm = TRUE)

kurtosis(transv$tEAPOS, na.rm = TRUE)

hist(transv$tEAPOS)

skewness(transvlog$logtEAPOS, na.rm = TRUE)

kurtosis(transvlog$logtEAPOS, na.rm = TRUE)

hist(transvlog$logtEAPOS)

# NEGATIVE EA:
skewness(transv$tEANEG, na.rm = TRUE)

kurtosis(transv$tEANEG, na.rm = TRUE)

hist(transv$tEANEG)

skewness(transvlog$logtEANEG, na.rm = TRUE)

kurtosis(transvlog$logtEANEG, na.rm = TRUE)

hist(transvlog$logtEANEG)

# CONSUMPTION CONTEXTS

# SOCIAL FACILITATION:
skewness(transv$tfacilitsoc, na.rm = TRUE)

kurtosis(transv$tfacilitsoc, na.rm = TRUE)

hist(transv$tfacilitsoc)

skewness(transvlog$logtfacilitsoc, na.rm = TRUE)

kurtosis(transvlog$logtfacilitsoc, na.rm = TRUE)

hist(transvlog$logtfacilitsoc)

# PEER GROUP ACCEPTANCE:
skewness(transv$taceptpares, na.rm = TRUE)

kurtosis(transv$taceptpares, na.rm = TRUE)

hist(transv$taceptpares)

skewness(transvlog$logtaceptpares, na.rm = TRUE)

kurtosis(transvlog$logtaceptpares, na.rm = TRUE)

hist(transvlog$logtaceptpares)

# PARENTAL CONTROL:
skewness(transv$tctrolparent, na.rm = TRUE)

kurtosis(transv$tctrolparent, na.rm = TRUE)

hist(transv$tctrolparent)

skewness(transvlog$logtctrolparent, na.rm = TRUE)

kurtosis(transvlog$logtctrolparent, na.rm = TRUE)

hist(transvlog$logtctrolparent)

# STRESS CONTROL:
skewness(transv$tctrolestres, na.rm = TRUE)

kurtosis(transv$tctrolestres, na.rm = TRUE)

hist(transv$tctrolestres)

skewness(transvlog$logtctrolestres, na.rm = TRUE)

kurtosis(transvlog$logtctrolestres, na.rm = TRUE)

hist(transvlog$logtctrolestres)
```



#### To assess  multicollinearity

To check for multicollinearity, we used the VIF (Variance Inflation Factor) values. We adopted the following criteria (Field et al., 2012):


If VIF values are less than 10 then that indicates there probably isn't cause for concern.
If the average of VIF values is not substantially greater than 1, then that also indicates that there's no cause for concern.
Tolerance below 0.1 and below 0.2 indicates a serious and a potential problem.


*Reference:* Field, A., Miles, J., & Field, Z. (2012). Discovering statistics using R. London: SAGE.


For assessing the assumption of no multicollinearity we used the vif() function, for the initial model:

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(car)
```


*WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES (OUTCOME)*


*MODEL INCLUDING SCORES (log transformed) OF EA AND CC IN THE CROSS-SECTIONAL SAMPLE*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
fitlogconsT<-lm(logtALCSEM ~ tsexo + tedad + tedadinicioalc + logtEAPOS + logtEANEG + logtfacilitsoc + logtaceptpares + logtctrolparent + logtctrolestres, data = transvlog)
```


*The VIF*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
vif(fitlogconsT)
```


*The mean VIF*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(vif(fitlogconsT))
```


*The tolerance*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
1/vif(fitlogconsT)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
detach("package:car", unload = TRUE)
```



Load the dataset (CROSS-SECTIONAL)

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# transv<-read.table("clipboard",header=TRUE, encoding = "Latin 1", dec=",", sep="\t") # CROSS-SECTIONAL
```


*HEAVY DRINKERS- CROSS-SECTIONAL*


*WOMEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
table(transv$tsexo)

transv[transv$tconsumosem>7 & transv$tsexo=="Wom",] # to select heavy drinkers that are only women. CDC criterium for heavy drinkers in women: consuming 8 drinks or more per week

transvwomheavy<-transv[transv$tconsumosem>7 & transv$tsexo=="Wom",] 

summary(transvwomheavy) 
```


*MEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
transv[transv$tconsumosem>14 & transv$tsexo=="Man",] # to select heavy drinkers that are only men. CDC criterium for heavy drinkers in men: consuming 15 drinks or more per week

transvmenheavy<-transv[transv$tconsumosem>14 & transv$tsexo=="Man",] 

summary(transvmenheavy) 
```


*NO HEAVY DRINKERS- CROSS-SECTIONAL*


*WOMEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
transv[transv$tconsumosem<8 & transv$tsexo=="Wom",] # to select NO heavy drinkers that are only women. CDC criterium for NO heavy drinkers in women: consuming LESS THAN 8 drinks per week

transvwomNOheavy<-transv[transv$tconsumosem<8 & transv$tsexo=="Wom",] 

summary(transvwomNOheavy) 
```


*MEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
transv[transv$tconsumosem<15 & transv$tsexo=="Man",] # to select NO heavy drinkers that are only men. CDC criterium for NO heavy drinkers in men: consuming LESS THAN 15 drinks per week

transvmenNOheavy<-transv[transv$tconsumosem<15 & transv$tsexo=="Man",] 

summary(transvmenNOheavy)
```


