---
title: "PaperAlcohol_LongitudinalSample.R"
author: "L. Cecilia López Steinmetz"
date: "17/2/2023"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

# R Code for:


# López Steinmetz LC, Leyes CA, Fong SB & Godoy JC. "Alcohol consumption, alcohol expectancies, and drinking contexts in young Argentinean college students before and during the COVID-19 pandemic: A one-year follow-up study"


### ABBREVIATIONS 


**Abbreviations:**
EA = Alcohol Expectancy (Expectativas hacia el alcohol)
CC = Consumption Contexts (Contextos de consumo de alcohol)


### LOAD THE DATASET: a) Longitudinal


#### a) Load the dataset: *LONGITUDINAL sample*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
dataset<-read.table("clipboard",header=TRUE, encoding = "Latin 1", dec=",", sep="\t") # LONGITUDINAL
```


#### To explore the data

#### a) *LONGITUDINAL sample*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(dplyr)
glimpse(dataset)  # here we can also check the type of the variables
summary(dataset)
```


*Check the type of the variables (LONGITUDINAL) and transform them (when necessary)*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
str(dataset)

dataset$sexo<-as.factor(dataset$sexo)
dataset$aunidalcsem<-as.numeric(dataset$aunidalcsem)
dataset$aALCSEM<-as.numeric(dataset$aALCSEM)
dataset$bunidalcsem<-as.numeric(dataset$bunidalcsem)
dataset$bALCSEM<-as.numeric(dataset$bALCSEM)
```


*To order categories*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
dataset$sexo <- factor(dataset$sexo, levels=c("mujer", "varon")) # to order categories

str(dataset)
```


### METHODS SECTION 


#### SKEWNESS, KURTOSIS, & MULTICOLLINEARITY


*To test Skewness and Kurtosis: See codes for supplementary files (at the end of these scripts)*


*To test for multicollinearity: See codes supplementary files (at the end of these scripts)*


#### DESCRIPTION OF THE SAMPLE


##### LONGITUDINAL 

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# Sex
table(dataset$sexo)

prop.table(table(dataset$sexo))*100


# Age
mean(dataset$aedad) # at the first measurement

sd(dataset$aedad) # at the first measurement


# Age of onset of alcohol consumption
mean(dataset$aedadinicioalc)

sd(dataset$aedadinicioalc)


# Weekly alcohol consumption (Units of alcohol): First measurement:
mean(dataset$aALCSEM)

sd(dataset$aALCSEM)


# Weekly alcohol consumption (Units of alcohol): Second measurement:
mean(dataset$bALCSEM, na.rm = TRUE)

sd(dataset$bALCSEM, na.rm = TRUE)

```



##### LOG TRANSFORMATIONS

*LONGITUDINAL SAMPLE*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
newdata <- dataset 
```


*Log transformatios*


*Time 1 (T1)*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES
newdata$logaALCSEM <- log(newdata$aALCSEM + 1) 

# ALCOHOL EXPECTANCIES

### POSITIVE EA. Time 1:
newdata$logaEAPOS <- log(newdata$aEAPOS + 1)

# NEGATIVE EA. Time 1:
newdata$logaEANEG <- log(newdata$aEANEG + 1)

# CONSUMPTION CONTEXTS

# SOCIAL FACILITATION. Time 1:
newdata$logafacilitsoc <- log(newdata$afacilitsoc + 1)

# PEER GROUP ACCEPTANCE. Time 1:
newdata$logaaceptpares <- log(newdata$aaceptpares + 1)

# PARENTAL CONTROL. Time 1:
newdata$logactrolparent <- log(newdata$actrolparent + 1)

# STRESS CONTROL. Time 1:
newdata$logactrolestres <- log(newdata$actrolestres + 1)
```


*Time 2 (T2)*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES
newdata$logbALCSEM <- log(newdata$bALCSEM + 1)

# ALCOHOL EXPECTANCIES

# POSITIVE EA. Time 2:
newdata$logbEAPOS <- log(newdata$bEAPOS + 1)

# NEGATIVE EA. Time 2:
newdata$logbEANEG <- log(newdata$bEANEG + 1)

# CONSUMPTION CONTEXTS

# SOCIAL FACILITATION. Time 2:
newdata$logbfacilitsoc <- log(newdata$bfacilitsoc + 1)

# PEER GROUP ACCEPTANCE. Time 2:
newdata$logbaceptpares <- log(newdata$baceptpares + 1)

# PARENTAL CONTROL. Time 2:
newdata$logbctrolparent <- log(newdata$bctrolparent + 1)

# STRESS CONTROL. Time 2:
newdata$logbctrolestres <- log(newdata$bctrolestres + 1)
```


### *LONGITUDINAL SAMPLE*


### *BASED ON MEASURES OF EA AND cc AT THE 1ST MEASUREMENT*


Preparing the data: based on the LOG TRANSFORMED VALUES

*To select the variables from the "newdata" dataset: FOR EA AND CC HERE WE USE THE SCORES AS MEASURED AT TIME 1*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(dplyr)
datasetlongt1<-select(newdata, code, sexo, aedad, aedadinicioalc, logaEAPOS, logaEANEG, logafacilitsoc, logaaceptpares, logactrolparent, logactrolestres, logaALCSEM, logbALCSEM)
```


*To convert the format of the dataframe into the long format*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(reshape)
longt1<-melt(datasetlongt1, id = c("code","sexo", "aedad", "aedadinicioalc", "logaEAPOS", "logaEANEG", "logafacilitsoc", "logaaceptpares", "logactrolparent", "logactrolestres"), measured = c("logaALCSEM","logbALCSEM"))
```


*To rename variables*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
names(longt1)<-c("participant","sexo", "aedad", "aedadinicioalc", "logaEAPOS", "logaEANEG", "logafacilitsoc", "logaaceptpares", "logactrolparent", "logactrolestres", "variable", "scores") # "variable" refers to the repeated-measures variable
```


*This creates a variable "period" in the dataframe longt1*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
longt1$period<-gl(2, 300, labels = c("first", "second")) # We created 2 sets of 300 scores, the labels option then specifies the names to attach to these 2 sets, which correspond to the levels of "period" (first measurement and follow-up).
```


*To make it clearer that there are two observations for each participant we have sorted the data by participant*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
longt1<-longt1[order(longt1$participant),]

longt1[1:6, ]

library(mice)

set.seed(123)

iniLONGT1 <- mice(longt1, maxit = 0)

predLONGT1 <- iniLONGT1$pred
predLONGT1
```


*Setting values of variables I'd like to leave out to 0 in the predictor matrix*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
predLONGT1[, c("variable", "period")] <- 0
predLONGT1

predLONGT1["scores", ] <- c(-2, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0) # The class variable (only one is allowed) is coded by a `-2'.

impLONGT1 <- mice(longt1, meth = c("", "", "", "", "", "", "", "", "", "", "", "2l.norm", ""), m = 10, pred = predLONGT1, maxit = 5, seed = 123) # The dependent variable is "scores" (alcohol consumption)

print(impLONGT1) 
impLONGT1$imp$scores # to see the imputed values
```


*Distribution of observed/imputed values*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
xyplot(impLONGT1, scores ~ logaEAPOS | .imp, pch = 20, cex=1.4)
xyplot(impLONGT1, scores ~ logaEANEG | .imp, pch = 20, cex=1.4)
xyplot(impLONGT1, scores ~ logafacilitsoc | .imp, pch = 20, cex=1.4)
xyplot(impLONGT1, scores ~ logaaceptpares | .imp, pch = 20, cex=1.4)
xyplot(impLONGT1, scores ~ logactrolparent | .imp, pch = 20, cex=1.4)
xyplot(impLONGT1, scores ~ logactrolestres | .imp, pch = 20, cex=1.4)
```


*Complete the dataset*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
myDatalongt1 <- complete(impLONGT1, 10) # WE IMPUTED THE 10TH DATASET
```


*To select the order in which the predictors will be included into the model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(leaps)

bestLONGT1<-regsubsets(scores ~ period + sexo + aedad + aedadinicioalc + logaEAPOS + logaEANEG + logafacilitsoc + logaaceptpares + logactrolparent + logactrolestres, data = myDatalongt1, nvmax = 11, method = "exhaustive")

summary(bestLONGT1, all.best = TRUE)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
plot(bestLONGT1, scale = "r2", main = "Best R^2") # R2 = 0.16
plot(bestLONGT1, scale = "adjr2", main = "Best Adjusted R^2") # Adj R2 = 0.15 
```


*MULTILEVEL ANALYSYS (MIXED EFFECTS)*


*Building the model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(nlme)

interceptOnlyT1 <-gls(scores ~ 1, data = myDatalongt1, method = "ML") # only the intercept

model1<-lme(scores ~ 1, random = ~1|participant, data = myDatalongt1, method = "ML", control = lmeControl(opt = "optim")) # the nested structure indicates repeated measures (within variable) # repeated measures nested within participants
anova(interceptOnlyT1, model1)  # to compare models
```


*To see the overall effect of each main effect we added them to the model one at a time*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1facsoc<-update(model1, .~. + logafacilitsoc)
summary(t1facsoc)  
anova(interceptOnlyT1, model1, t1facsoc) 
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1ctrlstress<-update(t1facsoc, .~. + logactrolestres)
summary(t1ctrlstress)  
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress) 
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1ctrlparental<-update(t1ctrlstress, .~. + logactrolparent)
summary(t1ctrlparental)  
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1edadinicio<-update(t1ctrlparental, .~. + aedadinicioalc)
summary(t1edadinicio) 
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1aceptpares<-update(t1edadinicio, .~. + logaaceptpares)
summary(t1aceptpares) 
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio, t1aceptpares)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1sexo<-update(t1aceptpares, .~. + sexo)
summary(t1sexo) 
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio, t1aceptpares, t1sexo)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1edad<-update(t1sexo, .~. + aedad)
summary(t1edad)  
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio, t1aceptpares, t1sexo, t1edad)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1period<-update(t1edad, .~. + period)
summary(t1period)  
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio, t1aceptpares, t1sexo, t1edad, t1period)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1eapos<-update(t1period, .~. + logaEAPOS)
summary(t1eapos)  
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio, t1aceptpares, t1sexo, t1edad, t1period, t1eapos)  
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t1eaneg<-update(t1eapos, .~. + logaEANEG)
summary(t1eaneg) 
anova(interceptOnlyT1, model1, t1facsoc, t1ctrlstress, t1ctrlparental, t1edadinicio, t1aceptpares, t1sexo, t1edad, t1period, t1eapos, t1eaneg)
```



```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
summary(t1aceptpares)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
intervals(t1aceptpares)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(pastecs)
by(myDatalongt1$scores, list(myDatalongt1$period), stat.desc, basic = FALSE)
by(myDatalongt1$scores, list(myDatalongt1$period, myDatalongt1$sexo), stat.desc, basic = FALSE)
```


*Calculating effect sizes*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(DSUR.noof)
```


We got effect sizes of meaningful predictors by executing: rcontrast(t,df)


rcontrast(t-value, DF) # this data is in: summary(t1aceptpares)

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# logafacilitsoc  
rcontrast(3.773315, 294)


# logactrolestres   
rcontrast(2.908588, 294)


# logactrolparent    
rcontrast(2.024686, 294)


# aedadinicioalc     
rcontrast(-2.092601, 294)


# logaaceptpares
rcontrast(-2.208698, 294)
```


### *LONGITUDINAL SAMPLE*


### *BASED ON MEASURES OF EA AND cc AT THE 2ND MEASUREMENT (or FOLLOW-UP)*


We have run the same analysis, but now considering the EA and CC scores as measured at time 2 (follow-up).


*Preparing the data: based on LOG TRANSFORMED VALUES*


*To select the variables from the "newdata" dataset: FOR EA AND CC HERE WE USE THE SCORES AS MEASURED AT TIME 2*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(dplyr)
datasetlongt2<-select(newdata, code, sexo, aedad, aedadinicioalc, logbEAPOS, logbEANEG, logbfacilitsoc, logbaceptpares, logbctrolparent, logbctrolestres, logaALCSEM, logbALCSEM)
```


*To convert the format of the dataframe into the long format*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(reshape)
longt2<-melt(datasetlongt2, id = c("code","sexo", "aedad", "aedadinicioalc", "logbEAPOS", "logbEANEG", "logbfacilitsoc", "logbaceptpares", "logbctrolparent", "logbctrolestres"), measured = c("logaALCSEM","logbALCSEM"))
```


*To rename variables*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
names(longt2)<-c("participant","sexo", "aedad", "aedadinicioalc", "logbEAPOS", "logbEANEG", "logbfacilitsoc", "logbaceptpares", "logbctrolparent", "logbctrolestres", "variable", "scores") # "variable" refers to the repeated-measures variable
```


*This creates a variable "period" in the dataframe longt2*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
longt2$period<-gl(2, 300, labels = c("first", "second")) # We created 2 sets of 300 scores, the labels option then specifies the names to attach to these 2 sets, which correspond to the levels of "period" (first measurement and follow-up).
```


*To make it clearer that there are two observations for each participant we have sorted the data by participant*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
longt2<-longt2[order(longt2$participant),]

longt2[1:6, ]
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(mice)
library(VIM)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# Missing data
p <- function(x) {sum(is.na(x))/length(x)*100}
apply(longt2, 2, p) 


library(mice)

set.seed(123)

iniLONGT2 <- mice(longt2, maxit = 0)

predLONGT2 <- iniLONGT2$pred
predLONGT2
```


*Setting values of variables I'd like to leave out to 0 in the predictor matrix*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
predLONGT2[, c("variable", "period")] <- 0
predLONGT2

predLONGT2["scores", ] <- c(-2, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0) # The class variable (only one is allowed) is coded by a `-2'.

impLONGT2 <- mice(longt2, meth = c("", "", "", "", "pmm", "pmm", "pmm", "pmm", "pmm", "pmm", "", "2l.norm", ""), m = 10, pred = predLONGT2, maxit = 5, seed = 123) # The dependent variable is "scores" (alcohol consumption)

print(impLONGT2) 
impLONGT2$imp$scores # to see the imputed values
```


*Distribution of oberserved/imputed values*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
xyplot(impLONGT2, scores ~ logbEAPOS | .imp, pch = 20, cex=1.4)
xyplot(impLONGT2, scores ~ logbEANEG | .imp, pch = 20, cex=1.4)
xyplot(impLONGT2, scores ~ logbfacilitsoc | .imp, pch = 20, cex=1.4)
xyplot(impLONGT2, scores ~ logbaceptpares | .imp, pch = 20, cex=1.4)
xyplot(impLONGT2, scores ~ logbctrolparent | .imp, pch = 20, cex=1.4)
xyplot(impLONGT2, scores ~ logbctrolestres | .imp, pch = 20, cex=1.4)
```


*Complete the dataset*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
myDatalongt2 <- complete(impLONGT2, 10) # WE IMPUTED THE 10TH DATASET
```


*To select the order in which the predictors will be included into the model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(leaps)

bestLONGT2<-regsubsets(scores ~ period + sexo + aedad + aedadinicioalc + logbEAPOS + logbEANEG + logbfacilitsoc + logbaceptpares + logbctrolparent + logbctrolestres, data = myDatalongt2, nvmax = 11, method = "exhaustive")

summary(bestLONGT2, all.best = TRUE)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
plot(bestLONGT2, scale = "r2", main = "Best R^2") # R2 = 0.14 
plot(bestLONGT2, scale = "adjr2", main = "Best Adjusted R^2") # Adj R2 = 0.13 
```


*MULTILEVEL ANALYSYS (MIXED EFFECTS)*


*Building the model*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(nlme)

interceptOnlyT2 <-gls(scores ~ 1, data = myDatalongt2, method = "ML") # only the intercept

modelfu2<-lme(scores ~ 1, random = ~1|participant, data = myDatalongt2, method = "ML", control = lmeControl(opt = "optim")) # the nested structure indicates repeated measures (within variable) # repeated measures nested within participants
anova(interceptOnlyT2, modelfu2)  # to compare models
```


*To see the overall effect of each main effect we added them to the model one at a time*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2facsoc<-update(modelfu2, .~. + logbfacilitsoc)
summary(t2facsoc)  
anova(interceptOnlyT2, modelfu2, t2facsoc) 
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2ctrlparental<-update(t2facsoc, .~. + logbctrolparent)
summary(t2ctrlparental)  
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental)  
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2ctrlstress<-update(t2ctrlparental, .~. + logbctrolestres)
summary(t2ctrlstress)  
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress) 
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2sexo<-update(t2ctrlstress, .~. + sexo)
summary(t2sexo) 
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2aceptpares<-update(t2sexo, .~. + logbaceptpares)
summary(t2aceptpares) 
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo, t2aceptpares)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2eaneg<-update(t2aceptpares, .~. + logbEANEG)
summary(t2eaneg) 
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo, t2aceptpares, t2eaneg)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2period<-update(t2eaneg, .~. + period)
summary(t2period)  
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo, t2aceptpares, t2eaneg, t2period) 
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2edadinicio<-update(t2period, .~. + aedadinicioalc)
summary(t2edadinicio) 
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo, t2aceptpares, t2eaneg, t2period, t2edadinicio)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2edad<-update(t2edadinicio, .~. + aedad)
summary(t2edad) 
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo, t2aceptpares, t2eaneg, t2period, t2edadinicio, t2edad)
```

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t2eapos<-update(t2edad, .~. + logbEAPOS)
summary(t2eapos)  
anova(interceptOnlyT2, modelfu2, t2facsoc, t2ctrlparental, t2ctrlstress, t2sexo, t2aceptpares, t2eaneg, t2period, t2edadinicio, t2edad, t2eapos)  
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
summary(t2ctrlparental)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
intervals(t2ctrlparental)
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(pastecs)
by(myDatalongt2$scores, list(myDatalongt2$period), stat.desc, basic = FALSE)
by(myDatalongt2$scores, list(myDatalongt2$period, myDatalongt2$sexo), stat.desc, basic = FALSE)
```


*Calculating effect sizes*

We got effect sizes of meaningful predictors by executing: rcontrast(t,df)

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# library(DSUR.noof)

# logbfacilitsoc   
rcontrast(3.910679, 298)


# logbctrolparent
rcontrast(3.748983, 298)
```





### SUPPLEMENTARY MATERIALS


#### To test Skewness and Kurtosis

#### & log transformation


*Testing Skewness and Kurtosis*


*Criteria:* range of acceptable values -1 to +1 for Skewness and -3 to +3 for Kurtosis (Brown, 2006) or near to.


*Reference:* Brown, T. A. (2006). Confirmatory factor analysis for applied research. New York: Guilford Press.


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(moments)
```


We have tested skewness and kurtosis in the original data (from which included NAs).
Then, we have tested skewness and kurtosis transforming the data by using the log transformation (natural logarithm).


#### *LONGITUDINAL SAMPLE*


*Time 1 (T1)*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES
skewness(newdata$aALCSEM)

kurtosis(newdata$aALCSEM)

hist(newdata$aALCSEM)

skewness(newdata$logaALCSEM)

kurtosis(newdata$logaALCSEM)

hist(newdata$logaALCSEM)

# ALCOHOL EXPECTANCIES

# POSITIVE EA. Time 1:
skewness(newdata$aEAPOS)

kurtosis(newdata$aEAPOS)

hist(newdata$aEAPOS)

skewness(newdata$logaEAPOS)

kurtosis(newdata$logaEAPOS)

hist(newdata$logaEAPOS)

# NEGATIVE EA. Time 1:
skewness(newdata$aEANEG)

kurtosis(newdata$aEANEG)

hist(newdata$aEANEG)

skewness(newdata$logaEANEG)

kurtosis(newdata$logaEANEG)

hist(newdata$logaEANEG)

# CONSUMPTION CONTEXTS

# SOCIAL FACILITATION. Time 1:
skewness(newdata$afacilitsoc)

kurtosis(newdata$afacilitsoc)

hist(newdata$afacilitsoc)

skewness(newdata$logafacilitsoc)

kurtosis(newdata$logafacilitsoc)

hist(newdata$logafacilitsoc)

# PEER GROUP ACCEPTANCE. Time 1:
skewness(newdata$aaceptpares)

kurtosis(newdata$aaceptpares)

hist(newdata$aaceptpares)

skewness(newdata$logaaceptpares)

kurtosis(newdata$logaaceptpares)

hist(newdata$logaaceptpares)

# PARENTAL CONTROL. Time 1:
skewness(newdata$actrolparent)

kurtosis(newdata$actrolparent)

hist(newdata$actrolparent)

skewness(newdata$logactrolparent)

kurtosis(newdata$logactrolparent)

hist(newdata$logactrolparent)

# STRESS CONTROL. Time 1:
skewness(newdata$actrolestres)

kurtosis(newdata$actrolestres)

hist(newdata$actrolestres)

skewness(newdata$logactrolestres)

kurtosis(newdata$logactrolestres)

hist(newdata$logactrolestres)
```



*Time 2 (T2)*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES
skewness(newdata$bALCSEM, na.rm = TRUE)

kurtosis(newdata$bALCSEM, na.rm = TRUE)

hist(newdata$bALCSEM)

skewness(newdata$logbALCSEM, na.rm = TRUE)

kurtosis(newdata$logbALCSEM, na.rm = TRUE)

hist(newdata$logbALCSEM)

# ALCOHOL EXPECTANCIES

# POSITIVE EA. Time 2:
skewness(newdata$bEAPOS, na.rm = TRUE)

kurtosis(newdata$bEAPOS, na.rm = TRUE)

hist(newdata$bEAPOS)

skewness(newdata$logbEAPOS, na.rm = TRUE)

kurtosis(newdata$logbEAPOS, na.rm = TRUE)

hist(newdata$logbEAPOS)

# NEGATIVE EA. Time 2:
skewness(newdata$bEANEG, na.rm = TRUE)

kurtosis(newdata$bEANEG, na.rm = TRUE)

hist(newdata$bEANEG)

skewness(newdata$logbEANEG, na.rm = TRUE)

kurtosis(newdata$logbEANEG, na.rm = TRUE)

hist(newdata$logbEANEG)

# CONSUMPTION CONTEXTS

# SOCIAL FACILITATION. Time 2:
skewness(newdata$bfacilitsoc, na.rm = TRUE)

kurtosis(newdata$bfacilitsoc, na.rm = TRUE)

hist(newdata$bfacilitsoc)

skewness(newdata$logbfacilitsoc, na.rm = TRUE)

kurtosis(newdata$logbfacilitsoc, na.rm = TRUE)

hist(newdata$logbfacilitsoc)

# PEER GROUP ACCEPTANCE. Time 2:
skewness(newdata$baceptpares, na.rm = TRUE)

kurtosis(newdata$baceptpares, na.rm = TRUE)

hist(newdata$baceptpares)

skewness(newdata$logbaceptpares, na.rm = TRUE)

kurtosis(newdata$logbaceptpares, na.rm = TRUE)

hist(newdata$logbaceptpares)

# PARENTAL CONTROL. Time 2:
skewness(newdata$bctrolparent, na.rm = TRUE)

kurtosis(newdata$bctrolparent, na.rm = TRUE)

hist(newdata$bctrolparent)

skewness(newdata$logbctrolparent, na.rm = TRUE)

kurtosis(newdata$logbctrolparent, na.rm = TRUE)

hist(newdata$logbctrolparent)

# STRESS CONTROL. Time 2:
skewness(newdata$bctrolestres, na.rm = TRUE)

kurtosis(newdata$bctrolestres, na.rm = TRUE)

hist(newdata$bctrolestres)

skewness(newdata$logbctrolestres, na.rm = TRUE)

kurtosis(newdata$logbctrolestres, na.rm = TRUE)

hist(newdata$logbctrolestres)
```




#### To assess  multicollinearity

To check for multicollinearity, we used the VIF (Variance Inflation Factor) values. We adopted the following criteria (Field et al., 2012):


If VIF values are less than 10 then that indicates there probably isn't cause for concern.
If the average of VIF values is not substantially greater than 1, then that also indicates that there's no cause for concern.
Tolerance below 0.1 and below 0.2 indicates a serious and a potential problem.


*Reference:* Field, A., Miles, J., & Field, Z. (2012). Discovering statistics using R. London: SAGE.


For assessing the assumption of no multicollinearity we used the vif() function, for the initial model:

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
library(car)
```


*WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES (OUTCOME)*

*MODEL INCLUDING SCORES (log transformed) OF EA AND CC DURING THE FIRST MEASUREMENT*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
fitlogconsA<-lm(logaALCSEM ~ sexo + aedad + aedadinicioalc + logaEAPOS + logaEANEG + logafacilitsoc + logaaceptpares + logactrolparent + logactrolestres, data = newdata)
```


*The VIF*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
vif(fitlogconsA)
```


*The mean VIF*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(vif(fitlogconsA))
```


*The tolerance*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
1/vif(fitlogconsA)
```


*WEEKLY CONSUMPTION OF ALCOHOLIC BEVERAGES (OUTCOME)*


*MODEL INCLUDING SCORES (log transformed) OF EA AND CC DURING THE SECOND MEASUREMENT*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
fitlogconsB<-lm(logbALCSEM ~ sexo + aedad + aedadinicioalc + logbEAPOS + logbEANEG + logbfacilitsoc + logbaceptpares + logbctrolparent + logbctrolestres, data = newdata)
```


*The VIF*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
vif(fitlogconsB)
```


*The mean VIF*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(vif(fitlogconsB))
```


*The tolerance*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
1/vif(fitlogconsB)
```



Load the dataset (LONGITUDINAL)

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
# dataset<-read.table("clipboard",header=TRUE, encoding = "Latin 1", dec=",", sep="\t") # LONGITUDINAL
```


```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
table(dataset$sex) # 210 women, 90 men
```


*HEAVY DRINKERS- LONGITUDINAL*


*WOMEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
dataset[dataset$aconsumosem>7 & dataset$sexo=="mujer",] # to select heavy drinkers that are only women. CDC criterium for heavy drinkers in women: consuming 8 drinks or more per week

longwomheavy<-dataset[dataset$aconsumosem>7 & dataset$sexo=="mujer",] 

summary(longwomheavy) 
```


*MEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
dataset[dataset$aconsumosem>14 & dataset$sexo=="varon",] # to select heavy drinkers that are only men. CDC criterium for heavy drinkers in men: consuming 15 drinks or more per week

longmenheavy<-dataset[dataset$aconsumosem>14 & dataset$sexo=="varon",] 

summary(longmenheavy) 
```


*NO HEAVY DRINKERS- LONGITUDINAL*


*WOMEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
dataset[dataset$aconsumosem<8 & dataset$sexo=="mujer",] # to select NO heavy drinkers that are only women. CDC criterium for NO heavy drinkers in women: consuming LESS THAN 8 drinks per week

longwomNOheavy<-dataset[dataset$aconsumosem<8 & dataset$sexo=="mujer",] 

summary(longwomNOheavy)
```


*MEN*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
dataset[dataset$aconsumosem<15 & dataset$sexo=="varon",] # to select NO heavy drinkers that are only men. CDC criterium for NO heavy drinkers in men: consuming LESS THAN 15 drinks per week

longmenNOheavy<-dataset[dataset$aconsumosem<15 & dataset$sexo=="varon",] 

summary(longmenNOheavy)
```


 

*Cases with missing data (NAs) in the follow-up:*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
zNA<-dataset[!complete.cases(dataset$bconsumosem), ] # both sexes with NAs in the variable weekly alcohol consumption: n = 181
zwomNA<-dataset[!complete.cases(dataset$bconsumosem) & dataset$sexo == "mujer", ] # women with NAs in the variable weekly alcohol consumption: n = 116
zmenNA<-dataset[!complete.cases(dataset$bconsumosem) & dataset$sexo == "varon", ] # men with NAs in the variable weekly alcohol consumption: n = 65
```


*Cases without missing data (NAs) in the follow-up*
```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
zCOMPL<-dataset[complete.cases(dataset$bconsumosem), ] # both sexes without NAs in the variable weekly alcohol consumption: n = 119
zwomCOMPL<-dataset[complete.cases(dataset$bconsumosem) & dataset$sexo == "mujer", ] # women without NAs in the variable weekly alcohol consumption: n = 94
zmenCOMPL<-dataset[complete.cases(dataset$bconsumosem) & dataset$sexo == "varon", ] # men without NAs in the variable weekly alcohol consumption: n = 25
```


*Mean age of participants without missing data (NAs) in the follow-up*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(zCOMPL$aedad) # both sexes

sd(zCOMPL$aedad) # both sexes


mean(zwomCOMPL$aedad) # women

sd(zwomCOMPL$aedad) # women


mean(zmenCOMPL$aedad) # men

sd(zmenCOMPL$aedad) # men
```


*Mean age of participants with missing data (NAs) in the follow-up*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(zNA$aedad) # both sexes

sd(zNA$aedad) # both sexes
 

mean(zwomNA$aedad) # women

sd(zwomNA$aedad) # women


mean(zmenNA$aedad) # men

sd(zmenNA$aedad) # men
```


*Mean weekly alcohol consumption (during the first measurement) among participants without missing data (NAs) in the follow-up*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(zCOMPL$aconsumosem) # both sexes

sd(zCOMPL$aconsumosem) # both sexes


mean(zwomCOMPL$aconsumosem) # women

sd(zwomCOMPL$aconsumosem) # women


mean(zmenCOMPL$aconsumosem) # men

sd(zmenCOMPL$aconsumosem) # men
```


*Mean weekly alcohol consumption (during the first measurement) among participants with missing data (NAs) in the follow-up*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
mean(zNA$aconsumosem) # both sexes

sd(zNA$aconsumosem) # both sexes


mean(zwomNA$aconsumosem) # women

sd(zwomNA$aconsumosem) # women


mean(zmenNA$aconsumosem) # men

sd(zmenNA$aconsumosem) # men
```


*Differences between means*


*Age: without vs with missing values in both sexes*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t.test(x = zCOMPL$aedad, y = zNA$aedad, paired = FALSE, alternative = "two.sided", mu = 0, conf.level = 0.95)
```


*Age: without vs with missing values in women*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t.test(x = zwomCOMPL$aedad, y = zwomNA$aedad, paired = FALSE, alternative = "two.sided", mu = 0, conf.level = 0.95)
```


*Age: without vs with missing values in men*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t.test(x = zmenCOMPL$aedad, y = zmenNA$aedad, paired = FALSE, alternative = "two.sided", mu = 0, conf.level = 0.95)
```


*Weekly alcohol consumption: without vs with missing values in both sexes*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t.test(x = zCOMPL$aconsumosem, y = zNA$aconsumosem, paired = FALSE, alternative = "two.sided", mu = 0, conf.level = 0.95)
```


*Weekly alcohol consumption: without vs with missing values in women*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t.test(x = zwomCOMPL$aconsumosem, y = zwomNA$aconsumosem, paired = FALSE, alternative = "two.sided", mu = 0, conf.level = 0.95)
```


*Weekly alcohol consumption: without vs with missing values in men*

```{r eval=TRUE, echo=TRUE, warning=TRUE, message=TRUE, error=TRUE}
t.test(x = zmenCOMPL$aconsumosem, y = zmenNA$aconsumosem, paired = FALSE, alternative = "two.sided", mu = 0, conf.level = 0.95)
```

